The role of Similarity in native spoken word recognition

### 2025-02-07 manuscript/manuscript.qmd
Minor styling changes

When some German speakers listened to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them misheard the line "I've got the power" as *Agatha Bauer*. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. Outrageous as these examples might sound, the reader may know a few cases in their own native language. This auditory illusion is common across languages and cultures and can feel quite real, often inevitable [@dembeck2015oberflachenubersetzung; @efimova2018homophonic]. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"). From a linguistic point of view, Soramimi is a particular case of homophonic translation: words or phrases that appear in the speech stream in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning [@gasparov2006semen]. Homophonic translation has mostly received attention as a perceptual curiosity, and as a literary figure, intentionally employed by translators to preserve the meter and "sound structure" (e.g., rhymes) of an original text in the language of translation [e.g., @levick2016translating; @pilshchikov2016semiotics]^[The rationale behind this translation approach is to "sacrifice[s] fidelity to the image rather than the melodiousness of verse" [@briusov1911ot]. For instance, Edith Grossman's translation of *El Quijote* [@cervantes] present a similar case, in which formal aspects of the original text in Spanish are favoured to some degree over equivalence to English, even sometimes keeping Spanish words from the original text in the English translation (e.g., *Señor*, *ínsula*).]. Little attention has been paid to homophonic translation as a psycholinguistic event.

In one of the few available studies concerning homophonic translation as a psycholinguistic phenomenon, @otake2007interlingual analysed 194 instances of Soramimi, broadcasted between 1992 and 2007 by the Japanese TV show *Soramimi hour*, in which examples of Soramimi were presented to the audience for comedy purposes. These instances consisted of homophonic translations of English song lyrics to words and phrases in Japanese. The vast majority of the reported instances were phrases (96%) and few were single words (4%). There was relatively high variability in the degree to which phonetic features of the presented English lyrics were preserved, with some resulting Japanese words or phrases sharing high overlap with the original English strings of sounds, and some sharing very little overlap. An analysis of the few instances of homophonic translations of single words revealed three main phonological processes that could explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /\textipa{kraI}/ to *kurai*, /\textipa{kuRai}/ [dark]), deletion (e.g., *go*, /\textipa{goU}/ to *go*, /go/ [go], and alternation (e.g., *low*, /\textipa{loU}/ to *rou*, /\textipa{Roo}/ [wax]). This suggests that the reported homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to the Japanese phoneme inventory and Japanese phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic]. The phonological processes described by @otake2007interlingual do not take place in a vaccum, but rather in the context of a lexicon. As words or phrases in the native lexicon are activated by their acoustic similarity with the non-native speech stream, they trigger a series of mechanisms that result in the selection of one of the candidate word or phrase in the native language. In the present work, we investigated the interplay between phonological similarity and one of the main mechanisms involved in the dynamics of lexical selection: phonological neighbourhood density. The phonological neighbourhood density of a given word-form refers to the number of words-forms that differ in a single phoneme from the former. For instance, the English word *cat* hbelongs to a dense phonological neighbourhood, with 50 English phonological neighbours (e.g., hat, bat, cap), while the word *charger* belongs to a sparse phonological neighbourhood, with six neighbours (e.g., charged,larger, charter, charmer). Phonological neighbourhood density affects lexical processing. For instance, words from dense neighbourhoods are recognized more slowly and less accuractely in the auditory modality than words from sparse neighbourhoods [e.g., @luce1998recognizing; @vitevitch1998words; @dufour2010phonological; @vitevitch2006clustering]. In this study, we examined how listeners may activate lexical representations in their native language when listening to an unfamiliar language.

Although homophonic translations do not necessarily take place with the preservation of meaning---which results in outrageous mistranslations like the aforementioned ones---homophonic translations sometimes may lead to correct translations. For example, imagine an English native listening to Dutch for the first time. This listener may encounter the word /\textipa{"pINgUIn}/ (*pinguin*), which sounds similar to its English translation /\textipa{"pENgwIn}/ (penguin). Given enough phonological similarity between the non-native word and its translation in the native language, a naïve listener may succeed at word comprehension when listening to an unfamiliar language. Such phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and (orthographic similarity) of the former to the equivalent in their native language.

@@

We further investigated the effect of phonological competitors. Even if the presented word and its target translation share high phonological similarity, participants may still provide incorrect translations if the presented word also shares high phonological similarity with other words in the native language. Words with denser phonological neighbourhoods are recognised more slowly and less accurately than words with sparser phonological neighbourhoods [@dufour2003lexical; @goldinger1989priming; @hamburger1996phonological; @luce1990similarity; @luce1998recognizing]. This is especially true if such neighbours share higher phonological similarity with the presented word, or are lexically more frequent than the target word [@luce1998recognizing]. Such interference effect of phonological neighbourhood density has also been reported across languages [@weber2004lexical]. We hypothesised that the size of the facilitatory role of phonological similarity would be inversely proportional to the amount of higher-frequency phonological neighbours of the presented word in the target language. In the case of non-native speech recognition, the presence of cross-linguistic pairs which are phonologically similar but differ in meaning (e.g., false friends) may act as distractors during lexical access, obstructing the selection of the appropriate target translation in the listener’s lexicon. For instance, @otwinowska2019more reported that false friends were disadvantaged relative to non-cognates by Polish second language learners, in contrast to cognates which were known better. It is therefore important to investigate the joint effect of both cognates and false friends when investigating the effect cross-linguistic phonological similarity has on word recognition in a foreign language. This could shed light on available strategies and challenges associated with the early stages of second language acquisition. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as *bottom* instead of as its correct translation *button*, due to combined effect of the close phonological similarity between *botón* and *bottom* along with the high lexical frequency of *bottom* relative to *button*. To test this prediction, we developed a lexical frequency-dependent measure of cross-language phonological neighbourhood density, in which a neighbour is counted only if it is higher frequency and is one phoneme apart from the presented word. If the phonological neighbourhood density of the target translation affects participants’ performance negatively, this would suggest that competitors in the native language affect recognition of non-native words during recognition of foreign speech. We conducted a series of three experiments to test these predictions.

@@

When some German speakers listened to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them misheard the line "I've got the power" as *Agatha Bauer*. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. Outrageous as these examples might sound, the reader may know a few cases in their own native language. This auditory illusion is common across languages and cultures and can feel quite real, often inevitable [@dembeck2015oberflachenubersetzung; @efimova2018homophonic]. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"). From a linguistic point of view, Soramimi is a particular case of homophonic translation: words or phrases that appear in the speech stream in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning [@gasparov2006semen]. Homophonic translation has mostly received attention as a perceptual curiosity, and as a literary figure, intentionally employed by translators to preserve the meter and "sound structure" (e.g., rhymes) of an original text in the language of translation [e.g., @levick2016translating; @pilshchikov2016semiotics]^[The rationale behind this translation approach is to "sacrifice[s] fidelity to the image rather than the melodiousness of verse" [@briusov1911ot]. For instance, Edith Grossman's translation of *El Quijote* [@cervantes] present a similar case, in which formal aspects of the original text in Spanish are favoured to some degree over equivalence to English, even sometimes keeping Spanish words from the original text in the English translation (e.g., *Señor*, *ínsula*).]. Little attention has been paid to homophonic translation as a psycholinguistic event.

In one of the few available studies concerning homophonic translation as a language processing phenomenon, @otake2007interlingual analysed 194 instances of Soramimi, broadcasted between 1992 and 2007 by the TV show *Soramimi hour*, in which examples of Soramimi were presented the the audience for comedy purposes. These instances consisted of homophonic translations of English song lyrics to words and phrases in Japanese. The vast majority of the reported instances were phrases (96%) and few were single words (4%). There was relatively high variability in the degree to which phonetic features of the presented English lyrics were preserved, with some resulting Japanese words or phrases sharing high overlap with the original English strings of sounds, and some sharing very little overlap. An analysis of the few instances of homophonic translations of single words revealed three main phonological processes that could explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /krʌɪ/ to *kurai*, /kuɾai/ [dark]), deletion (e.g., *go*, /goʊ/ to *go*, /go/ [go], and alternation (e.g., *low*, /loʊ/ to *rou*, /ɾoo/ [wax]). This suggests that the reported homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to the Japanese phoneme inventory and Japanese phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic].

The phonological processes described by @otake2007interlingual do not take place in a vaccum, but rather in the context of a lexicon. As words or phrases in the native lexicon are activated by their acoustic similarity with the non-native speech stream, they trigger a series of mechanisms that result in the selection of one of the candidate word or phrase in the native language. In the present work, we investigated the interplay between phonological similarity and one of the main mechanisms involved in the dynamics of lexical selection: phonological neighbourhood density. In particular, we examined how listeners may activate lexical representations in their native language when listening to an unfamiliar language. Although homophonic translations do not necessarily take place with the preservation of meaning---which results in outrageous mistranslations like the aforementioned ones---homophonic translations sometimes may lead to correct translations. For example, imagine an English native listening to Dutch for the first time. This listener may encounter the word /ˈpɪŋ.guɪn/ (*pinguin*), which sounds similar to its English translation /ˈpɛŋ.ɡwɪn/ (penguin). Given enough phonological similarity between the non-native word and its translation in the native language, a naïve listener may succeed at word comprehension when listening to an unfamiliar language. Such phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and orthographic similarity of the former to the equivalent in their native language.

When comparing the phonemic transcription of the French to that of its Spanish translation /ˈpweɾ.ta/ (*puerta*), one may be led to infer that these two words do not have much overlap. But one thing to note is that the voiced uvular fricative consonant /ʁ/ and the open-mid back rounded vowel /ɔ/, which do not exist in Spanish as phonemic categories, can be perceived as allophonic variations the native (Spanish) phonemes /ɾ/ and /o/, as described by assimilation models of phoneme perception [@best2001discrimination; @best1988examination]. Although this phonetic mismatch between the non-native word and its translation can have a noticeable toll on comprehension @cutler2004patterns, the Spanish listener may still be able to activate the phonological representation of /ˈpweɾ.ta/ (*puerta*) from hearing the word-form /ˈpɔʁt/ (*porte*), ultimately allowing them to access the correct semantic representation. Native prosody also constrains the occurrence of homophonic translations. For instance, @kentner2015rhythmic presented in the auditory modality French-native and German-native participants with English songs with lyrics. Homophonic translations followed language-specific segmentation strategies. Most homophonic translations reported by German-native participants resulted from inserting phrase and word boundaries before stressed syllables (following the more frequent trochaic pattern in German). On the contrary, most homophonic translations reported by French-native participants resulted from inserting phrase and word boundaries after stressed syllables, following the more frequent iambic pattern in French.

To explore the lexical mechanisms triggered by unfamiliar speech listening, we used a translation elicitation task in which participants listened to words from an unfamiliar language, and then tried to guess the words translations in their native language. We will henceforth refer to the auditory-presented words heard by participants on each trial as *presented words*, and the correct translation for the presented words as *target translations*. By manipulating the amount of phonological similarity between the presented words and their target translations, we explored listeners' reliance on phonological similarity to succeed in the task. Since participants were unfamiliar with the presented language, they could only use phonological similarity between the presented language and their native language to successfully translate the words. We therefore predicted that participants’ performance should increase when the translation pairs are phonologically more similar. We also predicted that there is a minimum threshold of phonological similarity to be sufficient for translation.

We further investigated the effect of phonological competitors. Even if the presented word and its target translation share high phonological similarity, participants may still provide incorrect translations if the presented word also shares high phonological similarity with other words in the native language. Words with denser phonological neighbourhoods are recognised more slowly and less accurately than words with sparser phonological neighbourhoods [@dufour2003lexical; @goldinger1989priming; @hamburger1996phonological; @luce1990similarity; @luce1998recognizing]. This is especially true if such neighbours share higher phonological similarity with the presented word, or are lexically more frequent than the target word [@luce1998recognizing]. Such interference effect of phonological neighbourhood density has also been reported across languages [@weber2004lexical]. We hypothesised that the size of the facilitatory role of phonological similarity would be inversely proportional to the amount of higher-frequency phonological neighbours of the presented word in the target language. In the case of non-native speech recognition, the presence of cross-linguistic pairs which are phonologically similar but differ in meaning (e.g., false friends) may act as distractors during lexical access, obstructing the selection of the appropriate target translation in the listener’s lexicon. For instance, @otwinowska2019more reported that false friends were disadvantaged relative to non-cognates by Polish second language learners, in contrast to cognates which were known better. It is therefore important to investigate the joint effect of both cognates and false friends when investigating the effect cross-linguistic phonological similarity has on word recognition in a foreign language. This could shed light on available strategies and challenges associated with the early stages of second language acquisition. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as *bottom* instead of as its correct translation *button*, due to combined effect of the close phonological similarity between *botón* and *bottom* along with the high lexical frequency of *bottom* relative to *button*. To test this prediction, we developed a lexical frequency-dependent measure of cross-language phonological neighbourhood density, in which a neighbour is counted only if it is higher frequency and is one phoneme apart from the presented word. If the phonological neighbourhood density of the target translation affects participants’ performance negatively, this would suggest that competitors in the native language affect recognition of non-native words during recognition of foreign speech. We conducted a series of three experiments to test these predictions.

@@

collection_dates <- participants_1 |>
  pull(date) |>
  range() |>
  format("%B %dth, %Y")

n_participants <- dplyr::n_distinct(participants_1)
n_participants_group <- table(participants_1$group)
participants_age <- participants_1 |>
  summarise(across(age, lst(mean, sd, min, max)))
n_female <- table(participants_1$gender)
```
We collected data from `r n_participants` British English-native adults living in United Kingdom (*Mean* = `r round(participants_age$age_mean, 2)` years, *SD* = `r round(participants_age$age_sd, 2)`, Range = `r round(participants_age$age_min, 2)`-`r round(participants_age$age_max, 2)`, `r n_female[1]` female). Data collection took place from `r collection_dates[1]` to `r collection_dates[2]`. Participants were recruited via Prolific (5£ compensation) and SONA (compensation in academic credits). Participants gave informed consent before providing any data. The study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was by the University of Oxford Medical Sciences Inter-Divisional Research Ethics Committee (IDREC) (R60939/RE007). Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection.

@@

tbl_data |>
  gt(groupname_col = "exp", rowname_col = "group") |>
  fmt_number(matches("age")) |>
  fmt_integer(matches("min|max")) |>
  cols_merge_uncert(age_mean, age_sd) |>
  cols_merge_range(age_min, age_max) |>
  tab_spanner("Age", matches("age")) |>
  cols_label(
    group = "Group",
    exp = "Experiment",
    n_excluded = "N",
    age_mean = "Mean ± SD ",
    age_min = "Range",
    l2_lst = "L2"
  ) |>
  tab_style(
    cell_text(style = "italic"),
    cells_column_labels()
  ) |>
  tab_style(
    cell_text(weight = "bold"),
    cells_column_spanners()
  ) |>
  tab_style(
    cell_text(align = "left"),
    list(cells_body(), cells_column_labels())
  ) |>
  tab_style(
    cell_text(
      weight = "bold",
      align = "center",
      style = "normal"
    ),
    cells_column_labels(c(l2_lst, n_excluded))
  ) |>
  tab_footnote(
    "Number of included participants (number of excluded participants.)",
    locations = cells_column_labels(n_excluded)
  )
```

@@

In Experiment 1, we investigated the extent to which the phonological similarity between translation equivalents is sufficient for successful word translation, in the absence of conceptual knowledge about the presented word. We tested two groups of monolingual British English-native adults in a translation task that involved words in Catalan or Spanish, two languages participants reported having no prior familiarity with. Participants benefited strongly from phonological similarity when the correct translation of the presented words in Catalan or Spanish had few English phonological neighbours with higher lexical frequency. This suggests that word-forms in an unfamiliar language have a strong potential to activate their translation equivalents in the native language, provided some phonological similarity between both words, and the absence of more frequent phonological neighbors.

>>>>

When some German speakers listen to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them mishear the line "I've got the power" as *Agatha Bauer*. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. Outrageous as these examples might sound, the reader may know a few cases in their own native language. This auditory illusion is common across languages and cultures and can feel quite real, often inevitable [@dembeck2015oberflachenubersetzung; @efimova2018homophonic]. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"). From a linguistic point of view, Soramimi is a particular case of homophonic translation: words or phrases that appear in the speech stream in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning [@gasparov2006semen]. Homophonic translation has mostly received attention as a perceptual curiosity, and as a literary figure, intentionally employed by translators to preserve the meter and "sound structure" (e.g., rhymes) of an original text in the language of translation [e.g., @levick2016translating; @pilshchikov2016semiotics]^[The rationale behind this translation approach is to "sacrifice[s] fidelity to the image rather than the melodiousness of verse" [@briusov1911ot]. For instance, Edith Grossman's translation of *El Quijote* [@cervantes] present a similar case, in which formal aspects of the original text in Spanish are favoured to some degree over equivalence to English, even sometimes keeping Spanish words from the original text in the English translation (e.g., *Señor*, *ínsula*).]. Little attention has been paid to homophonic translation from a language processing perspective.

One exception is the study by @otake2007interlingual, who analysed 194 instances of Soramimi, broadcasted between 1992 and 2007 by the Japanese TV show *Soramimi hour*, in which examples of Soramimi were presented to the audience for comedy purposes. These instances consisted of homophonic translations of English song lyrics to words and phrases in Japanese. The vast majority of the reported instances were phrases (96%) and a few, single words (4%). There was relatively high variability in the degree to which phonetic features of the presented English lyrics were preserved, with some resulting Japanese words or phrases sharing high overlap with the original English strings of sounds, and some sharing very little overlap. An analysis of the few instances of homophonic translations of single words revealed three main phonological processes that could explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /\textipa{kraI}/ to *kurai* /\textipa{kuRai}/ [dark]), deletion (e.g., *go* /\textipa{goU}/ to *go* /go/ [*Go*, a board game], and alternation (e.g., *low* /\textipa{loU}/ to *rou* /\textipa{Roo}/ [wax]). This suggests that the reported homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to the Japanese phoneme inventory and Japanese phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic].

The phonological processes described by @otake2007interlingual do not take place in a vaccum, but rather in the context of a lexicon. As words or phrases in the native lexicon are activated by their acoustic similarity with the non-native speech stream, they trigger a series of mechanisms that result in the selection of one of the candidate word or phrase in the native language. In the present work, we investigated the interplay between phonological similarity and one of the main mechanisms involved in the dynamics of lexical selection: phonological neighbourhood density. The phonological neighbourhood density of a given word-form refers to the number of words-forms that differ in a single phoneme from the it. For instance, the English word *cat* belongs to a dense phonological neighbourhood, with 50 English phonological neighbours (e.g., hat, bat, cap), while the word *charger* belongs to a sparse phonological neighbourhood, with six neighbours (e.g., charged, larger, charter, charmer) [@marian2012clearpond]. Phonological neighbourhood density affects lexical processing. For instance, words from dense neighbourhoods are recognized more slowly and less accuractely in the auditory modality than words from sparse neighbourhoods [@luce1998recognizing; @vitevitch1998words; e.g.,  @dufour2010phonological; @vitevitch2006clustering]. In this study, we examined how listeners may activate lexical representations in their native language when listening to an unfamiliar language.

Although homophonic translations rarely involve preservation of meaning, they sometimes lead to correct translations. For example, imagine an English native listening to Dutch for the first time. This listener may encounter the word /\textipa{"pINgUIn}/ (*pinguin*), which sounds similar to its English translation /\textipa{"pENgwIn}/ (penguin). Given enough phonological similarity between the non-native word and its translation in the native language, a naïve listener may succeed at word comprehension when listening to an unfamiliar language. Such phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and (orthographic similarity) of the former to the equivalent in their native language.

@@

We further investigated the effect of phonological competition. Even if the presented word and its target translation share high phonological similarity, participants may still provide incorrect translations if the presented word also shares high phonological similarity with other words in the native language. Words with denser phonological neighbourhoods are recognised more slowly and less accurately than words with sparser phonological neighbourhoods [@dufour2003lexical; @goldinger1989priming; @hamburger1996phonological; @luce1990similarity; @luce1998recognizing]. This is especially true if such neighbours share higher phonological similarity with the presented word, or are lexically more frequent than the target word [@luce1998recognizing]. Such interference effect of phonological neighbourhood density has also been reported across languages [@weber2004lexical]. We hypothesised that the size of the facilitatory role of phonological similarity would be inversely proportional to the amount of higher-frequency phonological neighbours of the presented word in the target language. In the case of non-native speech recognition, the presence of cross-linguistic pairs which are phonologically similar but differ in meaning (e.g., false friends) may act as distractors during lexical access, obstructing the selection of the appropriate target translation in the listener’s lexicon. For instance, @otwinowska2019more reported that false friends were disadvantaged relative to non-cognates by Polish second language learners, in contrast to cognates which were known better. It is therefore important to investigate the joint effect of both cognates and false friends when investigating the effect that cross-linguistic phonological similarity has on word recognition in a foreign language. This could shed light on available strategies and challenges associated with the early stages of second language acquisition. For instance, one might expect English participants to incorrectly translate the Spanish word *botón* as *bottom* instead of as its correct translation *button*, due to the combined effect of the close phonological similarity between *botón* and *bottom* along with the high lexical frequency of *bottom* relative to *button*. To test this prediction, we developed a lexical frequency-dependent measure of cross-language phonological neighbourhood density, in which a neighbour is counted only if it is higher frequency and is one phoneme apart from the presented word. If the phonological neighbourhood density of the target translation affects participants’ performance negatively, this would suggest that competitors in the native language affect recognition of non-native words during recognition of foreign speech. We conducted a series of three experiments to test these predictions.

@@

When some German speakers listen to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them mishear the line "I've got the power" as *Agatha Bauer*. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. Outrageous as these examples might sound, the reader may know a few cases in their own native language. This auditory illusion is common across languages and cultures and can feel quite real, often inevitable [@dembeck2015oberflachenubersetzung; @efimova2018homophonic]. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"). From a linguistic point of view, Soramimi is a particular case of homophonic translation: words or phrases that appear in the speech stream in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning [@gasparov2006semen]. Homophonic translation has mostly received attention as a perceptual curiosity, and as a literary figure, intentionally employed by translators to preserve the meter and "sound structure" (e.g., rhymes) of an original text in the language of translation [e.g., @levick2016translating; @pilshchikov2016semiotics]^[The rationale behind this translation approach is to "sacrifice[s] fidelity to the image rather than the melodiousness of verse" [@briusov1911ot]. For instance, Edith Grossman's translation of *El Quijote* [@cervantes] present a similar case, in which formal aspects of the original text in Spanish are favoured to some degree over equivalence to English, even sometimes keeping Spanish words from the original text in the English translation (e.g., *Señor*, *ínsula*).]. Little attention has been paid to homophonic translation from a language processing perspective.

One exception is the study by @otake2007interlingual, who analysed 194 instances of Soramimi, broadcasted between 1992 and 2007 by the TV show *Soramimi hour*, in which examples of Soramimi were presented the the audience for comedy purposes. These instances consisted of homophonic translations of English song lyrics to words and phrases in Japanese. The vast majority of the reported instances were phrases (96%) and a few, single words (4%). There was relatively high variability in the degree to which phonetic features of the presented English lyrics were preserved, with some resulting Japanese words or phrases sharing high overlap with the original English strings of sounds, and some sharing very little overlap. An analysis of the few instances of homophonic translations of single words revealed three main phonological processes that could explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /krʌɪ/ to *kurai*, /kuɾai/ [dark]), deletion (e.g., *go* /goʊ/ to *go* /go/ [*Go*, a board game], and alternation (e.g., *low* /loʊ/ to *rou* /ɾoo/ [wax]). This suggests that the reported homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to the Japanese phoneme inventory and Japanese phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic].

The phonological processes described by @otake2007interlingual do not take place in a vaccum, but rather in the context of a lexicon. As words or phrases in the native lexicon are activated by their acoustic similarity with the non-native speech stream, they trigger a series of mechanisms that result in the selection of one of the candidate word or phrase in the native language. In the present work, we investigated the interplay between phonological similarity and one of the main mechanisms involved in the dynamics of lexical selection: phonological neighbourhood density. The phonological neighbourhood density of a given word-form refers to the number of words-forms that differ in a single phoneme from the it. For instance, the English word *cat* belongs to a dense phonological neighbourhood, with 50 English phonological neighbours (e.g., hat, bat, cap), while the word *charger* belongs to a sparse phonological neighbourhood, with six neighbours (e.g., charged, larger, charter, charmer) [@marian2012clearpond]. Phonological neighbourhood density affects lexical processing. For instance, words from dense neighbourhoods are recognized more slowly and less accuractely in the auditory modality than words from sparse neighbourhoods [@luce1998recognizing; @vitevitch1998words; e.g.,  @dufour2010phonological; @vitevitch2006clustering]. In this study, we examined how listeners may activate lexical representations in their native language when listening to an unfamiliar language.

Although homophonic translations rarely involve preservation of meaning, they sometimes lead to correct translations. For example, imagine an English native listening to Dutch for the first time. This listener may encounter the word /\textipa{"pINgUIn}/ (*pinguin*), which sounds similar to its English translation /\textipa{"pENgwIn}/ (penguin). Given enough phonological similarity between the non-native word and its translation in the native language, a naïve listener may succeed at word comprehension when listening to an unfamiliar language. Such phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and (orthographic similarity) of the former to the equivalent in their native language.

We further investigated the effect of phonological competition. Even if the presented word and its target translation share high phonological similarity, participants may still provide incorrect translations if the presented word also shares high phonological similarity with other words in the native language. Words with denser phonological neighbourhoods are recognised more slowly and less accurately than words with sparser phonological neighbourhoods [@dufour2003lexical; @goldinger1989priming; @hamburger1996phonological; @luce1990similarity; @luce1998recognizing]. This is especially true if such neighbours share higher phonological similarity with the presented word, or are lexically more frequent than the target word [@luce1998recognizing]. Such interference effect of phonological neighbourhood density has also been reported across languages [@weber2004lexical]. We hypothesised that the size of the facilitatory role of phonological similarity would be inversely proportional to the amount of higher-frequency phonological neighbours of the presented word in the target language. In the case of non-native speech recognition, the presence of cross-linguistic pairs which are phonologically similar but differ in meaning (e.g., false friends) may act as distractors during lexical access, obstructing the selection of the appropriate target translation in the listener’s lexicon. For instance, @otwinowska2019more reported that false friends were disadvantaged relative to non-cognates by Polish second language learners, in contrast to cognates which were known better. It is therefore important to investigate the joint effect of both cognates and false friends when investigating the effect that cross-linguistic phonological similarity has on word recognition in a foreign language. This could shed light on available strategies and challenges associated with the early stages of second language acquisition. For instance, one might expect English participants to incorrectly translate the Spanish word *botón* as *bottom* instead of as its correct translation *button*, due to the combined effect of the close phonological similarity between *botón* and *bottom* along with the high lexical frequency of *bottom* relative to *button*. To test this prediction, we developed a lexical frequency-dependent measure of cross-language phonological neighbourhood density, in which a neighbour is counted only if it is higher frequency and is one phoneme apart from the presented word. If the phonological neighbourhood density of the target translation affects participants’ performance negatively, this would suggest that competitors in the native language affect recognition of non-native words during recognition of foreign speech. We conducted a series of three experiments to test these predictions.

In Experiment 1, we collected data from two groups of British English natives living in the United Kingdom. One group was presented with Catalan words, the other with Spanish words. We examined the extent to which participants were able to use the phonological similarity between the presented word (in Catalan or Spanish) and its target translation to provide accurate responses. In Experiment 2, we tested a group of (European) Spanish natives in the same task, who were presented with Catalan words. Catalan and Spanish are two Romance languages whose close typological distance is reflected in the fact that they share many cognates, where English is a Germanic language that shares considerably fewer cognates with Catalan and Spanish. By testing participants translating words from typologically close or distant languages, we expected to widen the range of the phonological similarity scores of the translation pairs involved in the experimental task, therefore allowing us to explore potential cross-language differences in participant’s performance. One unexpected finding was that participants in Experiments 1 and 2 were surprisingly good at translating a subset of words which had low phonological similarity with their correct translation. We were concerned that this may be caused by some prior knowledge of specific words by our participants, as some Spanish words are commonly seen in media or product labels, making them familiar even to monolingual speakers of English. In Experiment 3, we collected additional data from a new group of British English natives. The design was closely modelled after Experiment 1, except that after providing their response in each trial, participants reported whether they had previous knowledge of the presented word. In the final section of this article, we present analyses on the joint data sets of Experiments 1 to 3.
:::

@@

collection_dates <- participants_1 |>
  pull(date) |>
  range() |>
  format("%B %dth, %Y")
collection_dates <- gsub(" 0", " ", collection_dates)

n_participants <- dplyr::n_distinct(participants_1)
n_participants_group <- table(participants_1$group)
participants_age <- participants_1 |>
  summarise(across(age, lst(mean, sd, min, max)))
n_female <- table(participants_1$gender)
```
We collected data from `r n_participants` British English-native adults living in the United Kingdom (*Mean* = `r round(participants_age$age_mean, 2)` years, *SD* = `r round(participants_age$age_sd, 2)`, Range = `r round(participants_age$age_min, 2)`-`r round(participants_age$age_max, 2)`, `r n_female[1]` female). Data collection took place from `r collection_dates[1]` to `r collection_dates[2]`. Participants were recruited via Prolific (£5 compensation) and SONA (compensation in academic credits). Participants gave informed consent before providing any data. The study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was by the University of Oxford Medical Sciences Inter-Divisional Research Ethics Committee (IDREC) (R60939/RE007). Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection.

@@

tbl_data |>
  gt(groupname_col = "exp", rowname_col = "group") |>
  fmt_number(matches("age")) |>
  fmt_integer(matches("min|max")) |>
  cols_merge_uncert(age_mean, age_sd) |>
  cols_merge_range(age_min, age_max) |>
  tab_spanner("Age", matches("age")) |>
  cols_label(
    group = "Group",
    exp = "Experiment",
    n_excluded = "N",
    age_mean = "Mean ± SD ",
    age_min = "Range",
    l2_lst = "Second language"
  ) |>
  tab_style(
    cell_text(style = "italic"),
    cells_column_labels()
  ) |>
  tab_style(
    cell_text(weight = "bold"),
    cells_column_spanners()
  ) |>
  tab_style(
    cell_text(align = "left"),
    list(cells_body(), cells_column_labels())
  ) |>
  tab_style(
    cell_text(
      weight = "bold",
      align = "center",
      style = "normal"
    ),
    cells_column_labels(c(l2_lst, n_excluded))
  ) |>
  tab_footnote(
    "Number of included participants (number of excluded participants.)",
    locations = cells_column_labels(n_excluded)
  )
```

@@

In Experiment 1, we investigated the extent to which the phonological similarity between translation equivalents is sufficient for successful word translation, in the absence of conceptual knowledge about the presented word. We tested two groups of monolingual British English-native adults in a translation task that involved words in Catalan or Spanish, two languages participants reported having no prior familiarity with. Participants benefited strongly from phonological similarity when the correct translation of the presented words in Catalan or Spanish had few English phonological neighbours with higher lexical frequency. As the number of higher-frequency English phonological neighbours increased, both participants' accuracy and the impact of phonological similarity approached zero. This suggests that word-forms in an unfamiliar language have a strong potential to activate their translation equivalents in the native language, provided some phonological similarity between both words, and the absence of more frequent phonological neighbors.
